Negative and Fractional Types

### 2012-06-27 rational_ext.tex
constructions

\paragraph*{Booleans and Conditionals.} 
Given any combinator {{c : b <-> b}} we can construct a combinator called
{{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
the combinator {{c}} is used to transform the value of type~{{b}}. If the
boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
write down the combinator for {{if_c}} in terms of {{c}} as 
{{ dist (;) ((id (*) c) (+) id) (;) factor }}.

\noindent The diagram below shows the input value of type {{(1+1)*b}}
processed by the distribute operator {{dist}}, which converts it into a value
of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
combinator~{{c}} is applied to the value of type~{{b}}. The right branch
which corresponds to the boolean being {{false}} passes along the value of
type {{b}} unchanged.

\begin{center}
\scalebox{1.0}{
%%subcode-line{pdfimage}[diagrams/if_c.pdf]
\includegraphics{diagrams/thesis/cnot.pdf}
}
\end{center}

% We will be seeing many more such wiring diagrams in this paper and it is
% useful to note some conventions about them. Wires indicate a
% value that can exist in the program. Each wire, whenever possible, is
% annotated with its type and sometimes additional information to help clarify
% its role. When multiple wires run in parallel, it means that those values
% exist in the system at the same time, indicating pair types. When there is a
% disjunction, we put a {{+}} between the wires. 
% Combinators for distribution {{dist}} and factoring {{factor}}
% are represented as triangles with their operator symbols in them. Other
% triangles may be used and, in each case, types or labels will be used to
% clarify their roles. Finally, we don't draw boxes for combinators such as
% {{id}}, commutativity, and associativity, but instead just shuffle the wires
% as appropriate.

The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
negates its second argument if the first argument is {{true}}. This
gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
equivalent construction that is useful is {{else_{not} }} where we
negate the second argument only if the first is {{false}}.

Similarly, we can iterate the construction of {{if_c}} to check several
bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
checks two booleans and negates the result wire only if they are both
{{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
universal reversible gate. We can generalize this construction to
{{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
they are all {{true}}.

%%
\paragraph*{Cloning.}
Although cloning is generally not allowed in reversible languages, it is
possible at the cost of having additional constant inputs. For example,
consider the gate {{else_{not} }}. Generally, the gate maps
{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
on the cases in which the second input is {{true}}, we get that the gate maps
{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
i.e., the gate clones the first input. A circuit of {{n}} parallel
{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
{{true}} inputs in the process.  Let us call this construction
{{clone^n_{bool} }}.

>>>>

The constructions in this section provide a brief overview to writing
programs in {{Pi}}. More details maybe found in previous work
(Sec. 3.3 \cite{infeffects} and the constructions in \cite{rc2012}).

\paragraph*{Booleans, Conditionals and Cloning.} 
We use the type {{1+1}} to denote {{bool}}, with {{left ()}} as
{{true}} and {{right ()}} as {{false}}. The combinator
{{not:bool<->bool}} can be expressed by {{swap+}} at the type
{{bool}}. Further, given any two combinators {{c1:b1<->b2}} and
{{c2:b1<->b2}}, we can write the conditional combinator
{{if_{c1,c2}:bool*b1<->bool*b2}} textually as 
{{dist(;)((id(*)c1)(+)(id(*)c2))(;)factor}} and graphically as:

\dgm{if-c1-c2}

\noindent
If {{c1 v |--> v'}} and {{c2 v |--> v''}}, we can verify that the
conditional maps {{if_{c1,c2} (true, v) |--> (true, v')}} and
{{if_{c1,c2} (false, v) |--> (false, v'')}}.  Several useful
combinators follow from this basic construction:

\begin{enumerate}
\item The combinator {{cnot:bool*bool<->bool*bool}} that negates the
  second input if the first input (called the control wire) is
  {{true}} may be defined as {{if_{not,id} }}.
\item The universal reversible combinator, the Toffoli gate, may be
  defined as
  {{if_{cnot,id}:bool*(bool*bool)<->bool*(bool*bool)}}. This can be
  extended to test {{n}} control wires and we call this combinator
  {{cnot^n:bool^{n+1}<->bool^{n+1} }}.
\item The combinator {{if_{id,not}:bool*bool<->bool*bool}} has the
  property that it maps {{(x, true)|-->(x, x)}} i.e. if the second bit
  is fixed to {{true}} it clones the first bit. We can generalize this
  to clone {{n}} bits and we write
  {{clone^n:bool^n*bool^n<->bool^n*bool^n}}. The second argument
  {{bool^n}} must be supplied {{true}} values to correctly simulate
  cloning.
\end{enumerate}

%% Given any combinator {{c : b <-> b}} we can construct a combinator called
%% {{if_c : bool*b <->bool*b}} in terms of {{c}}, where {{if_c}} behaves like a
%% one-armed $\mathit{if}$-expression. If the supplied boolean is {{true}} then
%% the combinator {{c}} is used to transform the value of type~{{b}}. If the
%% boolean is {{false}}, then the value of type {{b}} remains unchanged. We can
%% write down the combinator for {{if_c}} in terms of {{c}} as 
%% {{ dist (;) ((id (*) c) (+) id) (;) factor }}.

%% \noindent The diagram below shows the input value of type {{(1+1)*b}}
%% processed by the distribute operator {{dist}}, which converts it into a value
%% of type {{(1*b)+(1*b)}}. In the {{left}} branch, which corresponds to the
%% case when the boolean is {{true}} (i.e. the value was {{left ()}}), the
%% combinator~{{c}} is applied to the value of type~{{b}}. The right branch
%% which corresponds to the boolean being {{false}} passes along the value of
%% type {{b}} unchanged.

%% \begin{center}
%% \scalebox{1.0}{
%% %%subcode-line{pdfimage}[diagrams/if_c.pdf]
%% \includegraphics{diagrams/thesis/cnot.pdf}
%% }
%% \end{center}

%% % We will be seeing many more such wiring diagrams in this paper and it is
%% % useful to note some conventions about them. Wires indicate a
%% % value that can exist in the program. Each wire, whenever possible, is
%% % annotated with its type and sometimes additional information to help clarify
%% % its role. When multiple wires run in parallel, it means that those values
%% % exist in the system at the same time, indicating pair types. When there is a
%% % disjunction, we put a {{+}} between the wires. 
%% % Combinators for distribution {{dist}} and factoring {{factor}}
%% % are represented as triangles with their operator symbols in them. Other
%% % triangles may be used and, in each case, types or labels will be used to
%% % clarify their roles. Finally, we don't draw boxes for combinators such as
%% % {{id}}, commutativity, and associativity, but instead just shuffle the wires
%% % as appropriate.

%% The combinator {{if_{not} }} has type {{bool*bool<->bool*bool}} and
%% negates its second argument if the first argument is {{true}}. This
%% gate {{if_{not} }} is often referred to as the {{cnot}} gate. An
%% equivalent construction that is useful is {{else_{not} }} where we
%% negate the second argument only if the first is {{false}}.

%% Similarly, we can iterate the construction of {{if_c}} to check several
%% bits. The gate {{if_{cnot} }}, which we may also write as {{if^2_{not} }},
%% checks two booleans and negates the result wire only if they are both
%% {{true}}. The gate {{if^2_{not} }} is well known as the Toffoli gate and is a
%% universal reversible gate. We can generalize this construction to
%% {{if^n_{not} }} which checks {{n}} bits and negates the result wire only if
%% they are all {{true}}.

%% %%
%% \paragraph*{Cloning.}
%% Although cloning is generally not allowed in reversible languages, it is
%% possible at the cost of having additional constant inputs. For example,
%% consider the gate {{else_{not} }}. Generally, the gate maps
%% {{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
%% on the cases in which the second input is {{true}}, we get that the gate maps
%% {{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
%% i.e., the gate clones the first input. A circuit of {{n}} parallel
%% {{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
%% {{true}} inputs in the process.  Let us call this construction
%% {{clone^n_{bool} }}.
